The Ex-Ante View of Recommender System Design

### 2019-09-16 paper/filter_bubbles.tex
updated rec eval section

We denote by $u_{in}$ individual $i$'s realized utility from consuming item $n$. By realized utility we In particular, we consider that the realized utility derived from a given item can be decomposed in the following manner:
\begin{align*}
u_{in}= v_{in} + \beta v_n
\end{align*}
where $v_{in}$ denotes an idiosyncratic component -- i.e. consumer $i$'s idiosyncratic taste for good $n$ --  and $v_{n}$, a common-value component. One can interpret $v_n$ as a measure of how much good $n$ is valued in society in general and, in a sense, $v_{in}$ denotes how $i$ diverges from this overall ranking. The scalar $\beta \in \mathbb{R}_{+}$ denotes the degree to which utilities are idiosyncratic to each individual or common across individuals. If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta \rightarrow \infty$, every individual has the same preferences. [include some discussion of what this implies about the predictability of a user's preferences]

@@

As a result, there is a need to define alternative recommendation evaluation metrics that result in recommender systems that are more useful for users. One recent alternative has been proposed such as serendipity
\cite{kotkov2016survey, kotkov2016challenges}. Ideally, serendipitous recommendations ``have the quality of being both unexpected and useful" \cite{maksai2015predicting}. \cite{iaquinta2010can} defines serendipitous recommendations as follows:
\begin{quote}
\textit{A serendipitous recommendation helps the
user to find a surprisingly interesting item that
she might not have otherwise discovered (or it
would have been really hard to discover). [...]
Serendipity cannot happen if the user already
knows what is recommended to her, because a
serendipitous happening is by definition something new. Thus the lower is the probability that user knows an item, the higher is the
probability that a specific item could result
in a serendipitous recommendation}.
\end{quote}

All approaches to serendipity to our knowledge, however, do not take into account the \textit{beliefs} that users have over the valuations for the items. For instance, \cite{vargas2011rank, kaminskas2014measuring} propose unexpectedness metrics that look at the dissimilarity of the recommended items to what the recommender already knows the user likes either via a content-based approach or a collaborative-based approach. \cite{adamopoulos2015unexpectedness} use an expected utility approach but measure unexpectedness in terms of deviations from the ``expected consideration set" of a user and do not explicitly incorporate user beliefs.

We propose a definition of serendipity based on \textit{unexpectedness} and \textit{usefulness} which have have clear definitions in the context of our model of user decision-making. Users have \textit{beliefs} over the valuations for the items and thus \textit{unexpected} recommendations are given by items that consumers have low expected utility for, either due to the fact that their mean belief is low or the degree of uncertainty is high. Recall from Findings \ref{finding_local_consumption} and \ref{finding_diversity_welfare_corr} that, particularly if users are risk-averse, they will consume content in increasingly narrow portions of the product space and this may be detrimental to their realized welfare. As a result, identifying the portions of the product space where users may have significant uncertainty and utilizing recommendation to reduce this uncertainty can be welfare-enhancing for consumers. This motivates a definition of \textit{useful} recommendations as recommendations that give consumers information that would \textit{change the item they consume relative to no recommendation}. Combining these, serendipitious recommendations are recommendations that provide users with information about \textit{unexpected} items that end up being \textit{useful} by changing consumption choices to increase the realized utility of a user.

>>>>

We denote by $u_{in}$ individual $i$'s realized utility from consuming item $n$. In particular, we consider that the realized utility derived from a given item can be decomposed in the following manner:
\begin{align*}
u_{in}= v_{in} + \beta v_n
\end{align*}
where $v_{in}$ denotes an idiosyncratic component -- i.e. consumer $i$'s idiosyncratic taste for good $n$ --  and $v_{n}$, a common-value component. One can interpret $v_n$ as a measure of how much good $n$ is valued in society in general and, in a sense, $v_{in}$ denotes how $i$ diverges from this overall ranking. The scalar $\beta \in \mathbb{R}_{+}$ denotes the degree to which utilities are idiosyncratic to each individual or common across individuals. If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta \rightarrow \infty$, every individual has the same preferences. [include some discussion of what this implies about the predictability of a user's preferences]

@@

As a result, several alternative recomendation evaluation metrics, such as serendipity \cite{kotkov2016survey}, coverage \cite{ge2010beyond}, and novelty \cite{vargas2011rank}, have been proposed in order to develop recommender systems that produce recommendations that are more useful for users. We use our model to motivate an alternative recommendation evaluation metric that provides an alternative perspective on the serendipity metric proposed in the literature. Serendipitious recommendations ``have the quality of being both unexpected and useful" \cite{maksai2015predicting}. \cite{iaquinta2010can} defines serendipitous recommendations as follows:
\begin{quote}
\textit{A serendipitous recommendation helps the
user to find a surprisingly interesting item that
she might not have otherwise discovered (or it
would have been really hard to discover). [...]
Serendipity cannot happen if the user already
knows what is recommended to her, because a
serendipitous happening is by definition something new. Thus the lower is the probability that user knows an item, the higher is the
probability that a specific item could result
in a serendipitous recommendation}.
\end{quote}

Our approach to serendipity comes from the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without recommendation. This approach stems from the idea that the value of recommendation is the marginal gain in expected utility that a user receives from recommendation.\footnote{This notion is similar to the ``value of information" notion utilized in the literature in economics on the pricing of information goods \cite{bergemann2018design}.} This intuitively should come from providing information about products that, without additional information, users assign a low likelihood to being the best product currently available and so would not consume them without recommendation.

This differs from other approaches in the literature since it depends on the beliefs of the users and what item the user would consume without recommendation. For instance, \cite{vargas2011rank, kaminskas2014measuring} propose unexpectedness metrics that look at the dissimilarity of the recommended items to what the recommender already knows the user likes either via a content-based approach or a collaborative-based approach. This depends only on the resulting item-set that the users get recommended and not necessarily user beliefs or the change in user action. \cite{adamopoulos2015unexpectedness} use an expected utility approach but use this to measure unexpectedness in terms of deviations from the ``expected consideration set" of a user and do not explicitly incorporate user beliefs.

Our model allows us to give precise definitions for what it means for a recommendation to be \textit{unexpected} and \textit{useful}. Unexpected recommendations are given by recommended items that, given their beliefs, users have low expected utility from consuming, irregardless of what the true realized utility is. Recall from Findings \ref{finding_local_consumption} and \ref{finding_diversity_welfare_corr} that, particularly if users are risk-averse, they will consume content in increasingly narrow portions of the product space and this may be detrimental to their realized welfare. As a result, identifying the portions of the product space where users may have significant uncertainty and utilizing recommendation to reduce this uncertainty can be welfare-enhancing for consumers. This is especially the case if this reduction of uncertainty leads users to explore under-explored portions of the product space where true realized utility is high, but expected utility, given beliefs, is low before recommendation.

Useful recommendations are recommendations that give users information that would change the item they consume relative to no recommendation and lead to an increase in expected utility. In the previous example, recommending \textit{John Wick: Chapter Two} is not useful precisely because the expected utility before and after recommendation is the same since it does not change behavior. The most useful recommendations, or those that have the highest utility gain, are recommendations that induce users to consume unexpected goods.
